---
title: "2019-03-18 Files (and Lists) in Environment"
author: ''
date: '2019-03-18'
draft: false
output:
  blogdown::html_page:
    toc: true
    # toc_float: true error in this theme https://github.com/rstudio/blogdown/issues/58
    #smart: true
slug: files-list-management
categories:
  - Rtutorial
tags:
  - readxl
  - purrr
  - forloop
  - lists
image:
  caption: ''
  focal_point: ''
summary: "Write & read multiple `.xlsx/.csv/.Rdata` files to/from R environment"   
---

**GOAL**: Below are a few different ways to write & read multiple `.xlsx/.csv/.Rdata` files to/from R environment - while retaining the same name(s). <!--more-->




```{r global_options, include=FALSE}
# I am not executing bc it for the blog post
knitr::opts_chunk$set(
  eval = FALSE, echo = TRUE,
  tidy = FALSE,
  results = "hide",
  message = TRUE,
  warning = FALSE, fig.show = "hide", fig.align = "center", fig.width = 8, fig.height = 8
)
# load libraries
library(dplyr)
library(purrr)
library(readr)
library(tidyr)
library(readxl)
```




## 0) PRELIMINARY STEPS

#### Use some pre-loaded R datasets & save as DFs
```{r}
# data() 
mtcars <- as.data.frame(mtcars)
iris <- as.data.frame(iris)
orange <- as.data.frame(Orange)
titanic <- as.data.frame(Titanic)
Orange <- as.data.frame(Orange)
OrchardSprays <- as.data.frame(OrchardSprays)
airquality <- as.data.frame(airquality)
airmiles <- as.data.frame(airmiles)
```


#### Create a list `list` with names (2 ways to name elements of list)
```{r}
# 1.a) Create a list of n data frames
list_dfs <- list(mtcars, iris ,  orange, titanic)
list_dfs[1]

# 1.b) Give names the data frames
names(list_dfs) <- c("mtcars","iris", "orange" , "titanic") 

# OR -----

# 2) Create a list of n data frames WHILE giving them a  name
list_dfs_N <- list(mtcars = mtcars, iris = iris, orange = orange, titanic = titanic)
list_dfs_N[[1]]
```


#### Create Output DIR `file.path` in a proj folder (e.g. "./static/post/...")
```{r}
# Create Output Dir... remember final"/"
Outdf2csv <- file.path(".", "static","post", "df2csv/")
dir.create(Outdf2csv)

# Dir_pcr <- file.path(".", "static","post", "pcr/")
# dir.create(Dir_pcr)
```

--- 

## TASK 1) WRITE ?? (MAINTAINING NAMES)
 
### OPTION (1.1a) WRITE list of Envir objects as **.csv** using `ls` + `for`
**NOTE**
 
* Not ~~`length(list_loop_DF)`~~ (lenght of the list), but `seq_along(list_loop_DF)` -> *CORRECT*! (Generates a sequence long as the list) 
* `utils:write.table(x, file = "")` where
	+ x = `list_loop_DF[[i]]`
	+ file = `paste0(Outdf2csv, names(list_loop_DF[i]), ".csv")`


#### List of Envir objects --> many csv files 
```{r loop}
# Create a list of NAMED dataframes
list_loop_DF <- list(airquality = airquality, airmiles = airmiles)

# Write a .csv file with each
for (i in seq_along(list_loop_DF)) {
  # Outdf2csv <- if (!dir.exists("./zzz_purrr/Output/")) {
  #   dir.create(file.path("./zzz_purrr/Output/"))
  # } else {
  #   print("Dir already exists!")
  # }
  write.csv(x = list_loop_DF[[i]], 
  			 file = paste0(Outdf2csv, names(list_loop_DF[i]), ".csv"))
}


# or -----
for (i in seq_along(list_loop_DF)) {
  # Outdf2csv <- if (!dir.exists("./zzz_purrr/Output/")) {
  #   dir.create(file.path("./zzz_purrr/Output/"))
  # } else {
  #   print("Dir already exists!")
  # }
  readr::write_csv(x = list_loop_DF[[i]], 
  			 path = paste0(Outdf2csv, names(list_loop_DF[i]), ".csv"))
}

```

### OPTION (1.1b) WRITE list of Envir objects --> as **.Rdata** files using `ls` + `for`

> **NOTE** `save (list= ...., file = ...)` -> *list instead of obj name*!  

```{r}
# Create Output Dir... remember final'/'
OutAPI <- file.path(".", "static","post", "SendGrid_API/")
# dir.create(OutAPI)

# List of files starting  with <REcip_>
files_Recip <- ls( pattern="^Recip_"# "^xx" = "starts with  xx"
						 # all.files = FALSE,
						 # full.names = FALSE, recursive = FALSE,
						 # ignore.case = FALSE, include.dirs = FALSE, no.. = FALSE)
)   

# ====== LOOP for as many times as there are items in the listing (= LENGHT)
for (i in 1:length(files_Recip)) {
	# at each round  
	save (list = (files_Recip[i]) ,  # ... take the name of an item from the list 
			file = paste0(OutAPI, files_Recip[i], ".Rdata"))
}

```


### OPTION (1.2) WRITE list of Envir objects as **.csv** using `lapply`  (within a function) 

**NOTE** I already created above landing Output dir `Outdf2csv` - becomes:
```{r lapply }
# Create a list of NAMED dataframes
list_lapply_DF <- list(mtcars = mtcars, titanic = titanic ) # 

# Write the function with arguments (DFlist)
Func_list_lapply <- function(list_lapply_DF) { # optional arg2 (Outdf2csv)
	lapply(1:length(list_lapply_DF), 
			 function(i) write.csv(list_lapply_DF[[i]], 
			 							 file = paste0(Outdf2csv, 
			 							 				  names(list_lapply_DF[i]), ".csv"), 
			 							 row.names = FALSE
			 )
	)
}

# Call the function
Func_list_lapply(list_lapply_DF)
```

--- 

### OPTION (1.3) WRITE list of Envir objects as **.csv** using `purrr:walk`  (not ~~map~~ )
**NOTE** Writing a file to a disk is considered to be a side-effect (we are not changing our data, only save it) so we must use `purrr:walk`

> Specifically, we are using `purrr::walk2(.x, .y, .f, ...` where 
	`.x` and `.y` are vectors of the same length
	`.f` is a 2-argument function 
	
```{r purr}
library(purrr)
library(tidyverse)

# Create a list of NAMED dataframes
list_purrr_DF <- list(Orange = Orange, OrchardSprays = OrchardSprays ) # 

# Set the Output Dir to an object - done before... 
  # Outdf2csv <- if (!dir.exists("./zzz_purrr/Output/")) {
  #   dir.create(file.path("./zzz_purrr/Output/"))
  # } else {
  #   print("Dir already exists!")
  # }

# ... & combine it with the complete file path
path <- file.path(paste0(Outdf2csv, names(list_purrr_DF), ".csv"))

# run the function
purrr::walk2(list_purrr_DF, path, write.csv)
```

---


## OPTION (2) CLEAN && WRITE as **.csv**to disk `group_by + group_map + purrr::group_map`
[Luis Verde Arregoitia's](https://luisdva.github.io/rstats/export-iteratively/)

In this case, the five new files (one for each bat family) will end up in the working directory.
**NOTE** We use `group_by` and `group_map` to create a grouped tibble and `apply` functions to each group. 
+group_map returns a list, so we can use paste0 to create a path for each file to be written, including a custom prefix. 

```{r}
# read csv from web
batRecs <- read.csv("https://raw.githubusercontent.com/luisDVA/codeluis/master/batRecords.csv",
  stringsAsFactors = FALSE
)
# preview how many files we should be ending up with
batRecs %>% count(family)

# drop na,
batRecs %>% na.omit() %>%
  # split to create a list of data frames for each group,
  group_by(family) %>%
  # then map to apply functions to each list element. I
  group_map(~ distinct(.x, decimal_latitude, decimal_longitude, .keep_all = TRUE), keep = TRUE) %>%
  # walk because write.csv returns nothing and creates the csv file as a side effect
  walk(~ .x %>% 
  	  	write_csv(path = paste0(Outdf2csv, "nov1_", unique(.x$family), ".csv"), 
  	  				 append = F))
```

--- 

--- 

## TASK 2)  Reverse topic: READ into Environment 

### OPTION (2.1) READ list of *.csv into R environment `list.files` + `for` loop {`custom function`}

Here, I want to read all .csv files into R and create a DF for each (keeping the original csv file name
minus "spaces and ".csv".


```{r , eval=FALSE, results="hide"}
# path
path <- file.path(".", "static","post", "df2csv/")
# list of files 
list_2read <-  list.files(path = path, 
						# pattern = c('^Send','.csv$' ), 
                	full.names = F,  
						all.files = FALSE,  # def (= only visible)
                  recursive = FALSE,  # def  (= no inside sub-dir )
                  ignore.case = TRUE, # (= pattern-matching be case-insensitive)
                  include.dirs = FALSE, # def (subdir names NOT be included in recursive listings)
                  no.. = FALSE) %>% # def (both "." and ".." be excluded also from non-recursive listings) 
            sort(decreasing = FALSE) 

# CSV_read FUNCTION
for (file in list_2read) {
  order <- which( # Give the TRUE indices of a logical object
    strsplit(file, "")[[1]] == "."
  ) # Split the elements of a character vector "file" to the matches to substring split
  
  assign( # Assign a value to a name in an environment.
    gsub(" ", "", substr(file, 1, order - 1)), # x = a variable name
    # read.csv(paste(path,file,sep=""), stringsAsFactors = F) # value = to be assigned to x
    read_csv(paste(path, file, sep = ""))
  )
}
```


### OPTION (2.2) READ list of *.csv into R environment `list.files` + `stringr`+ `map`

Elegant version of above 

```{r , eval=FALSE, results="hide"}
# path
path <- file.path(".", "static","post", "df2csv/")
# list of files 
list_2read <-  list.files(path = path, 
						# pattern = c('^Send','.csv$' ), 
                	full.names = F,  
						all.files = FALSE,  # def (= only visible)
                  recursive = FALSE,  # def  (= no inside sub-dir )
                  ignore.case = TRUE, # (= pattern-matching be case-insensitive)
                  include.dirs = FALSE, # def (subdir names NOT be included in recursive listings)
                  no.. = FALSE) %>% # def (both "." and ".." be excluded also from non-recursive listings) 
            sort(decreasing = FALSE) 

# select only 2
list_2read %>% 
	.[stringr::str_detect(., "air")] -> csv_files

 
# Load everything into the Global Environment
csv_files %>%
  purrr::map(function(file_name){ # iterate through each file name
  # Give a name to :	
  assign(x = stringr::str_remove(file_name, ".csv"), # x = file, but without ".csv")
         value = read_csv(paste0(path, file_name)), # value -> it will take in env
         envir = .GlobalEnv) # (where)
})
```




--- 

## TASK 3) Render multiple .Rmd files saving outputs in separate DIR using `base::list.files + purrr::walk`

This is the case in which I have several `.Rmd` files in the working directory and I want to render them all in one go. 
**NOTE** The output types were specifies in the .Rmd files themselves like so : 
 
> uses {base::list.files + purrr::walk}

 

```{r}
#======  DIRECTORIES PATHS  
root_DIR <- fs::path_abs(".")
reports_DIR <- fs::path_abs("./reports") # output reports 

# ======== LIST THE FILES I NEED TO RENDER   
filesRmd <- list.files(path = ".", 
							  pattern=".Rmd$",
							  all.files = FALSE,  # def (= only visible)
							  full.names = TRUE,  # I NEED dir name prepended
							  recursive = FALSE,  # def  (= no inside sub-dir )
							  ignore.case = TRUE, # (= pattern-matching be case-insensitive)
							  include.dirs = FALSE, # def (subdirectory names NOT be included in recursive listings)
							  no.. = FALSE)   # def (both "." and ".." be excluded also from non-recursive listings) 
 

# Execute them them in the environment
purrr::walk(filesRmd, rmarkdown::render,
	output_file = NULL, #  If using NULL then the output filename will be based on filename for the input file
	output_format = "all", # because I specify it in the .Rmd files (Html & pdf) 
	output_dir = reports_DIR,
	envir = new.env(), # The environment in which the code chunks are to be evaluated during knitting
	encoding = "UTF-8"
)


```


---

## ACKNOWLEDGMENTS

+ [More operations](https://luisdva.github.io/rstats/export-iteratively/)
+ [Following this post](https://lsru.github.io/tv_course/TD_purrr_solution.html)
+ Prof Claus Wilke has a [Useful tutorial](https://serialmentor.com/blog/2016/6/13/reading-and-combining-many-tidy-data-files-in-R)
+ [Martin Chan's post](https://martinctc.github.io/blog/vignette-write-and-read-multiple-excel-files-with-purrr/)
